The dynamics of starvation and recovery

### 2017-02-09 Manuscript/arxiv_version/tmp.tex
new yield constants

\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{margin=1in}

\usepackage[pdftex]{graphicx}

%\usepackage{amssymb,amsfonts,amsmath}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage[font=large]{caption}
\usepackage{subfigure}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{array}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
\usepackage{rotating}
\usepackage{color}
%\usepackage{lineno}
\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
\usepackage{authblk}
\usepackage{multicol}

\bibliographystyle{ieeetr}

@@

\title{The dynamics of starvation and recovery}%: Eco-evolutionary feedbacks}
\author[1,2]{Justin D. Yeakel\thanks{To whom correspondence should be addressed: jdyeakel@gmail.com}}
\author[2]{Christopher P. Kempes}
\author[2,3]{Sidney Redner}
\affil[1]{School of Natural Sciences, UC Merced, Merced, CA}
\affil[2]{The Santa Fe Institute, Santa Fe, NM}
\affil[3]{Department of Physics, Boston University, Boston, MA}

@@

\date{}

@@

\noindent {\bf Abstract} \\ %250 words
The eco-evolutionary dynamics of species is fundamentally linked to the energetic constraints of its constituent individuals.  
Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments.  
To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality.  
As a function of the transition rates between these replete and undernourished states that are determined by the presence or absence of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state.  
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics subject to these constraints can approach the cyclic regime but are typically driven to a steady state.  
Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized.  
Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations.  
Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percent body fat.  
For body masses $\lesssim 10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $\gtrsim 10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.

\begin{multicols}{2}

\noindent {\bf Introduction} \\ 
The behavioral ecology of all organisms is influenced by the energetic state of individuals, which directly influences how they invest reserves in uncertain environments.  
Such behaviors are generally manifested as tradeoffs between investing in somatic maintenance and growth, or allocating energy towards reproduction~\cite{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.  
The timing of these behaviors responds to selective pressure, as the choice of the investment impacts future fitness~\cite{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.  
The influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to repeated delays or shifts in reproduction over the course of an organism's life.

@@

In this work, we adopt an alternative approach in which we \emph{explicitly} account for resource limitation and the subsequent effects of starvation.
Namely, only individuals with sufficient energetic reserves can reproduce.
Such a constraint leads to reproductive time lags due to some members of the population going hungry and then recovering.  
Additionally, we incorporate the idea that reproduction is strongly constrained allometrically \cite{Kempes:2012hy}, and is not generally linearly related to resource density.  
As we shall show, these constraints influence the ensuing population dynamics in dramatic ways.
\\

\noindent {\bf Nutritional state-structured model (NSM)}\\
We begin by defining a minimal Nutritional State-structured population Model (NSM), where the consumer population is partitioned into two states: (a) an energetically replete (full) state $F$, where the consumer reproduces at a constant rate $\lambda$ and does not die from starvation, and (b) an energetically deficient (hungry) state $H$, where the consumer does not reproduce but dies by starvation at rate $\mu$.  
The underlying resource $R$ evolves by logistic growth with an intrinsic growth rate $\alpha$ and a carrying capacity equal to one.  
Consumers transition from the full state $F$ to the hungry state $H$ at a rate $\sigma$---the starvation rate---and also in proportion to the absence of resources $(1-R)$.  Conversely, consumers recover from state $H$ to state $F$ at rate $\rho$ and in proportion to $R$. 
Resources are also eaten by the consumers---at rate $\rho$ by hungry consumers and at rate $\beta<\rho$ by full consumers.  
This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.  
The NSM represents a fundamental extension of the idealized starving random walk model of foraging, which focuses on resource depletion, to include reproduction and resource replenishment~\cite{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}.

@@

Whereas the relation between consumer growth rate $\lambda$ and the starvation rate $\sigma$ defines an absolute bound of biological feasibility---the TC bifurcation---$\sigma$ also determines the sensitivity of the consumer population to changes in resource density.  
When $\sigma\gg\lambda$, the steady-state population density is small, thereby increasing the risk of stochastic extinction.  
On the other hand, as $\sigma$ decreases, the system will ultimately be poised either near the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).  
If the recovery rate $\rho$ is sufficiently small, the TC bifurcation is reached and the resource eventually is eliminated.  
If $\rho$ exceeds a threshold value, cyclic dynamics will develop as the Hopf bifurcation is approached.
\\

\noindent {\bf Role of allometry} \\ 
While there are no a priori constraints on the parameters in the NSM, most
organisms correspond to restricted portions of the parameter space.  Here we
use allometric scaling relations to constrain the covariation of rates in a
principled and biologically meaningful manner.  Allometric scaling relations
highlight common constraints and average trends across large ranges in body
size and species diversity. Many of these relations can be derived from a
small set of assumptions and below we describe a framework to determine the
covariation of timescales and rates across the range of mammals for each of
the key parameters of our model (cf.~\cite{Yodzis:1992hg}).  We are thereby
able to define the regime of dynamics occupied by the entire class of mammals
along with the key differences between the largest and smallest mammals.

@@

@@

@@

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$, to a
reduced-mass hungry state where reproduction is impossible.  For starving individuals we assume that an organism must meet its maintenance requirements using the digestion of existing mass as the sole energy source.
This assumption implies the following simple metabolic balance 
\begin{eqnarray}
\dot{m}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
or
\begin{eqnarray}
\dot{m}=-\frac{a^{\prime}}{M^{1-\eta}}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$, the energy required to
synthesis a unit of biomass \cite{hou}. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-a^{\prime}t/M^{1-\eta}}.
\end{eqnarray}
The time scale for starvation is
given by the time it takes $m(t)$ to reach $\epsilon_{\sigma} M$, which gives
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{M^{1-\eta}}{a^{\prime}}\ln\left(\epsilon_{\sigma}\right).
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/M^{1-\eta}\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Fig. \ref{fig:gvs}), which is important for the dynamics that we
later discuss.

@@

Although the rate equations \eqref{eq:system} are general, here we focus on
parameterizations for terrestrial-bound endotherms, specifically mammals,
which range from a minimum of $M\approx1$g (the Etruscan shrew
\emph{Suncus etruscus}) to a maximum of $M\approx10^7$g (the late Eocene
to early Miocene Indricotheriinae).  Investigating other classes of organisms
would simply involve altering the metabolic exponents and scalings associate
with $\epsilon$. Moreover, we emphasize that our allometric equations
describe mean relationships, and do not account for the (sometimes
considerable) variance associated with individual species.
\\

\noindent {\bf  Stabilizing effects of allometric constraints} \\
As the allometric derivations of the NSM rate laws reveal, starvation and recovery rates are not independent parameters, and the biologically relevant portion of the phase space shown in Fig.~\ref{fig:fp} is constrained via covarying parameters.  
Given the parameters of terrestrial endotherms, we find that the starvation rate $\sigma$ and the recovery rate $\rho$ are constrained to lie within a small window of potential values (Fig.~\ref{fig:hopf}) for the known range of body sizes $M$. 
We thus find that the dynamics for all mammalian body sizes is confined to the steady-state regime of the NSM and that limit-cycle behavior is precluded.
Moreover, for larger $M$, the distance to the Hopf bifurcation increases, while uncertainty in allometric parameters (20\% variation around the mean; Fig.~\ref{fig:hopf}) results in little qualitative difference in the distance to the the Hopf bifurcation. 
These results suggest that small mammals are more prone to population oscillations---both stable limit cycles and transient cycles---than mammals with larger body size.  
Thus our NSM model predicts that population cycles should be less common for larger species and more common for smaller species, particularly in environments where resources are limiting.

It should be noted that previous studies have used allometric constraints to explain the periodicity of cyclic populations \cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period $\propto M^{0.25}$.  
However this relation seems to hold only for some species \cite{Hendriks:2012fc}, and potential drivers range from predator and/or prey lifespans to competitive dynamics~\cite{Kendall:1999iy,Hogstedt:2005cr}.
Statistically significant support for the existence of population cycles among mammals is predominantly based on time series for small mammals~\cite{Kendall:1998hl}, in agreement with our predictions of more pronounced transient dynamics, given how close these points are to the Hopf bifurcation.  
On the other hand, the longer gestational times and the increased difficulty in measurements, precludes obtaining similar-quality data for larger organisms.
\vspace{6mm}

\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the population to the hungry state.  
In this state reproduction is absent, thus increasing the likelihood of extinction.  From the perspective of population survival, it is the rate of starvation relative to the rate of recovery that determines the long-term dynamics of the various species (Fig.~\ref{fig:fp}). 
We therefore examine the competing effects of cyclic dynamics vs.\ changes in steady state density on extinction risk as a function of $\sigma$ and $\rho$.  
To this end, we computed the probability of extinction, where we define extinction as a population trajectory falling below one tenth of the allometrically constrained steady state at any time between $t=10^5$ and $t \leq 10^8$.  
This procedure is repeated for 1000 replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an organism of $M=100$ grams.
In each replicate the initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random variable that is uniformly distributed in $[0,2]$.  
By allowing the rate of starvation to vary, we assessed extinction risk across a range of values for $\sigma$ and $\rho$ between ca.\ $10^{-6}$ to $10^{-1}$. 
As expected, higher rates of extinction correlate with both high values of $\sigma$ if $\rho$ is small, and high values of $\rho$ if $\sigma$ is small.  
For low values of $\sigma$ and high values of $\rho$, the increased extinction risk results from transient cycles with larger amplitudes as the system nears the Hopf bifurcation (Fig.~\ref{fig:ext}).  
For high values of $\sigma$ and low values of $\rho$, higher extinction risk arises because of the decrease in the steady state consumer population density (Figs. \ref{fig:fp}\emph{B}, \ref{fig:ext}).  
This interplay creates an `extinction refuge', such that for a constrained range of $\sigma$ and $\rho$, extinction probabilities are minimized.

We find that the allometrically constrained values of $\sigma$ and $\rho$ fall squarely within the extinction refuge (Fig. \ref{fig:ext}, white point). 
These values are close enough to the Hopf bifurcation to avoid low steady state densities, and far enough away to avoid large-amplitude transient cycles. 
The fact that allometric values of $\sigma$ and $\rho$ fall within this relatively small window supports the possibility that a selective mechanism has constrained the physiological conditions that drive starvation and recovery rates within populations.  
Such a mechanism would select for organism physiology that generates appropriate $\sigma$ and $\rho$ values that serve to minimize extinction risk.  
This selection could occur via the tuning of body fat percentages, metabolic rates, and biomass maintenance efficiencies.
To summarize, our finding that the allometrically-determined parameters fall within this low extinction probability region suggests that the NSM dynamics may both drive---and constrain---natural animal populations.\\

\noindent {\bf Dynamic and energetic barriers to body size} \\
Metabolite transport constraints are widely thought to place strict boundaries on biological scaling~\cite{Brown:1993p708,West:1997cg,Brown:2004wq} and thereby lead to specific predictions on the minimum possible body size for organisms~\cite{West:2002ud}.  
Above this bound, a number of energetic and evolutionary mechanisms have been explored to assess the costs and benefits associated with larger body masses, particularly for mammals.  
One important such example is the \emph{fasting endurance hypothesis}, which contends that larger body size, with consequent lower metabolic rates and increased ability to maintain more endogenous energetic reserves, may buffer organisms against environmental fluctuations in resource availability~\cite{Millar:1990p923}. 
Over evolutionary time, terrestrial mammalian lineages show a significant trend towards larger body size (known as Cope's rule)~\cite{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br}, and it is thought that within-lineage drivers generate selection towards an optimal upper bound of roughly $10^7$ grams~\cite{Alroy:1998p1594}, a value that is likely limited by higher extinction risk for large taxa over longer timescales~\cite{Clauset:2009fh}.  
These trends are thought to be driven by a combination of climate change and niche availability~\cite{Saarinen:2014br}; however the underpinning energetic costs and benefits of larger body sizes, and how they influence dynamics over ecological timescales, have not been explored.  
We argue that the NSM provides a suitable framework to explore these issues.

@@

To examine whether the NSM could provide such a mechanism, we begin by noting that a theoretical upper bound on mammalian body size is given by $\epsilon_\sigma=0$, where mammals are entirely composed of metabolic reserves, and this occurs at $M=8.3\times 10^8$, or $4.5\times$ the mass of a blue whale.
Next we examine to what extent a more realistic upper bound to body mass may serve as an evolutionary attractor, thus providing a suitable within-linneage mechanism for Cope's rule. 
We directly assess the susceptibility of an otherwise homogeneous population to invasion by a mutated subset of the population (denoted by ${}^\prime$) where individuals have a modified proportion of body fat $M^\prime=M(1+\chi)$ where $\chi \in [-1,1]$, thus altering the rates of starvation $\sigma(M^\prime)$, recovery $\rho(M^\prime)$, and maintenance $\beta(M^\prime)$.  
There is no internal fixed point corresponding to a state where both original residents and invaders coexist (except for the trivial state $\chi=0$).  
To assess the susceptibility to invasion as a function of the invader mass, we determine which consumer has a higher steady-state density for a given value of $\chi$.  
We find that for $1\leq M<8.43\times 10^6$g, having additional body fat ($\chi > 0$) results in a higher steady-state invader population density ($H^{\prime *}+F^{\prime *}>H^*+F^*$).  
Thus the invader has an intrinsic advantage over the resident population.  
However, for $M>8.43\times 10^6$, leaner individuals ($\chi < 0$) have advantageous steady state densities.

>>>>

\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
%\usepackage[pdftex]{graphicx}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage[english]{babel} 
\usepackage[latin1]{inputenc} 
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{float}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{stmaryrd} 
\usepackage{anyfontsize}

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{lineno}
%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

%\bibliographystyle{ieeetr}

@@

\author{Justin D. Yeakel} \affiliation{School of Natural Sciences, University
  of California, Merced, Merced, CA 95340, USA}

\author{Christopher P. Kempes} \affiliation{The Santa Fe Institute, 1399 Hyde
  Park Road, Santa Fe, NM 87501, USA}

\author{Sidney Redner} \affiliation{The Santa Fe Institute, 1399 Hyde Park
  Road, Santa Fe, NM 87501, USA}

\title{The dynamics of starvation and recovery}%: Eco-evolutionary feedbacks}

@@

\begin{abstract}

The eco-evolutionary dynamics of species is fundamentally linked to the
energetic constraints of its constituent individuals.  Of particular
importance are the tradeoffs between reproduction and the dynamics of
starvation and recovery in resource-limited environments.  To elucidate the
consequences of this tradeoff, we introduce a minimal nutritional
state-structured model that incorporates two classes of consumer:
nutritionally replete consumers that reproduce, and undernourished,
non-reproducing consumers that are susceptible to mortality.  As a function
of the transition rates between these replete and undernourished states that
are determined by the presence or absence of resources, the consumer
populations can either undergo cyclic dynamics or reach a steady state.  We
obtain strong constraints on starvation and recovery rates by deriving
allometric scaling relationships and find that population dynamics subject to
these constraints can approach the cyclic regime but are typically driven to
a steady state.  Moreover, we find that these rates fall within a `refuge' in
parameter space, where the probability of extinction of the consumer
population is minimized.  Thus we identify a potential mechanism that may
both drive and constrain the dynamics of animal populations.  Our model
provides a natural framework that predicts maximum body size for mammals by
determining the relative stability of an otherwise homogeneous population to
a mutant population with altered percent body fat.  For body masses
$\lesssim 10^7$g, individuals with increased energetic reserves can invade
resident populations, and vice versa for body mass $\gtrsim 10^7$g, thus
providing a principled mechanism for a within-lineage driver of Cope's rule.

\end{abstract}

%\begin{multicols}{2}

@@

\section*{Introduction}
The behavioral ecology of all organisms is influenced by the energetic state
of individuals, which directly influences how they invest reserves in
uncertain environments.  Such behaviors are generally manifested as tradeoffs
between investing in somatic maintenance and growth, or allocating energy
towards reproduction~\cite{Martin:1987dl,Kirk:1997cc,Kempes:2012hy}.  The
timing of these behaviors responds to selective pressure, as the choice of
the investment impacts future
fitness~\cite{Mangel:1988uaa,Mangel:2014kz,Yeakel:2013hi}.  The influence of
resource limitation on an organism's ability to maintain its nutritional
stores may lead to repeated delays or shifts in reproduction over the course
of an organism's life.

@@

In this work, we adopt an alternative approach in which we \emph{explicitly}
account for resource limitation and the subsequent effects of starvation.
Namely, only individuals with sufficient energetic reserves can reproduce.
Such a constraint leads to reproductive time lags due to some members of the
population going hungry and then recovering.  Additionally, we incorporate
the idea that reproduction is strongly constrained allometrically
\cite{Kempes:2012hy}, and is not generally linearly related to resource
density.  As we shall show, these constraints influence the ensuing
population dynamics in dramatic ways.

\section*{Nutritional state-structured model (NSM)}
We begin by defining a minimal Nutritional State-structured population Model (NSM), where the consumer population is partitioned into two states: (a) an energetically replete (full) state $F$, where the consumer reproduces at a constant rate $\lambda$ and does not die from starvation, and (b) an energetically deficient (hungry) state $H$, where the consumer does not reproduce but dies by starvation at rate $\mu$.  
The underlying resource $R$ evolves by logistic growth with an intrinsic growth rate $\alpha$ and a carrying capacity equal to one.  
Consumers transition from the full state $F$ to the hungry state $H$ at a rate $\sigma$---the starvation rate---and also in proportion to the absence of resources $(1-R)$.  Conversely, consumers recover from state $H$ to state $F$ at rate $\rho$ and in proportion to $R$. 
Resources are also eaten by the consumers---at rate $\rho$ by hungry consumers and at rate $\beta<\rho$ by full consumers.  
This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.  
The NSM represents a fundamental extension of the idealized starving random walk model of foraging, which focuses on resource depletion, to include reproduction and resource replenishment~\cite{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.75\textwidth]{fig_FixedPoint.pdf}
\caption{\small The transcritical (dashed) and Hopf bifurcation (solid) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
  bifurcation conditions separate parameter space into infeasible, cyclic,
  and steady state dynamic regimes.  The color gradient shows the steady
  state densities for (\emph{A}) the resource $R^*$ and the (\emph{B}) energetically
  replete consumers $F^*$, (warmer colors denote higher densities).  Steady
  state densities for the energetically deficient consumers $H^*$ (not shown)
  scale with those for $F^*$.  }
\label{fig:fp}
\end{figure*}

@@

Whereas the relation between consumer growth rate $\lambda$ and the
starvation rate $\sigma$ defines an absolute bound of biological
feasibility---the TC bifurcation---$\sigma$ also determines the sensitivity
of the consumer population to changes in resource density.  When
$\sigma\gg\lambda$, the steady-state population density is small, thereby
increasing the risk of stochastic extinction.  On the other hand, as $\sigma$
decreases, the system will ultimately be poised either near the TC or the
Hopf bifurcation (Fig.~\ref{fig:fp}).  If the recovery rate $\rho$ is
sufficiently small, the TC bifurcation is reached and the resource eventually
is eliminated.  If $\rho$ exceeds a threshold value, cyclic dynamics will
develop as the Hopf bifurcation is approached.

\section*{Role of allometry}
While there are no a priori constraints on the parameters in the NSM, most
organisms correspond to restricted portions of the parameter space.  Here we
use allometric scaling relations to constrain the covariation of rates in a
principled and biologically meaningful manner.  Allometric scaling relations
highlight common constraints and average trends across large ranges in body
size and species diversity. Many of these relations can be derived from a
small set of assumptions and below we describe a framework to determine the
covariation of timescales and rates across the range of mammals for each of
the key parameters of our model (cf.~\cite{Yodzis:1992hg}).  We are thereby
able to define the regime of dynamics occupied by the entire class of mammals
along with the key differences between the largest and smallest mammals.

@@

\begin{figure}
\includegraphics[width=0.375\textwidth]{Growth-trajectory-diagram.pdf}
\caption{\small The growth trajectory over absolute time of an individual
  organism as a function of body mass.  Initial growth follows the black
  trajectory to an energetically replete reproductive adult mass
  $m=\epsilon_\lambda M$ which we assume is 95\% asymptotic mass $M$.
  Starvation follows the red trajectory to
  $m = \epsilon_\sigma \epsilon_\lambda M$, and recovery follows the green
  growth trajectory to the replete adult mass. Alternatively, death from
  starvation follows the blue trajectory to
  $m=\epsilon_\mu \epsilon_\lambda M$.}
\label{fig:growth}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.45\textwidth]{fig_Rates.pdf}
\caption{\small Allometrically constrained starvation rate $\sigma$ (red) and
  recovery rate $\rho$ (green) relative to the reproductive rate $\lambda$
  (orange) as a function of body mass.  The rate of starvation is greater
  than the rate of reproduction for all realized terrestrial endotherm body
  sizes.  Mean values $\pm 20\%$ variation are shown by the shaded region for
  each rate.  }
\label{fig:gvs}
\end{figure}

@@

To determine the starvation rate, $\sigma$, we are interested in the time
required for an organism to go from a mature adult that reproduces at rate
$\lambda$, to a
reduced-mass hungry state where reproduction is impossible.  For starving individuals we assume that an organism must meet its maintenance requirements using the digestion of existing mass as the sole energy source.
This assumption implies the following simple metabolic balance 
\begin{eqnarray}
\dot{m}E_{m}^{\prime}=-B_{m}m
\end{eqnarray}
or
\begin{eqnarray}
\dot{m}=-\frac{a^{\prime}}{M^{1-\eta}}m
\end{eqnarray}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing
body mass which differs from $E_{m}$, the energy required to
synthesis a unit of biomass \cite{hou}. Given the replete mass, $M$, of an organism, the
above energy balance prescribes the mass trajectory of a non-consuming
organism:
\begin{eqnarray}
\label{mt}
m\left(t\right)=Me^{-a^{\prime}t/M^{1-\eta}}.
\end{eqnarray}
The time scale for starvation is
given by the time it takes $m(t)$ to reach $\epsilon_{\sigma} M$, which gives
\begin{equation}
\label{eq:sigma}
t_{\sigma}=-\frac{M^{1-\eta}}{a^{\prime}}\ln\left(\epsilon_{\sigma}\right).
\end{equation}
The starvation rate is then $\sigma=1/t_{\sigma}$, which scales with
replete-state mass as $1/M^{1-\eta}\ln\left(1-f_{0}M^{\gamma}/M\right)$.  An important
feature is that $\sigma$ does not have a simple scaling dependence on
$\lambda$ (Fig.~\ref{fig:gvs}), which is important for the dynamics that we
later discuss.

@@

Although the rate equations \eqref{eq:system} are general, here we focus on
parameterizations for terrestrial-bound endotherms, specifically mammals,
which range from a minimum of $M\approx1$g (the Etruscan shrew
\emph{Suncus etruscus}) to a maximum of $M\approx10^7$g (the late Eocene
to early Miocene Indricotheriinae).  Investigating other classes of organisms
would simply involve altering the metabolic exponents and scalings associate
with $\epsilon$. Moreover, we emphasize that our allometric equations
describe mean relationships, and do not account for the (sometimes
considerable) variance associated with individual species.

\section*{Stabilizing effects of allometric constraints}
As the allometric derivations of the NSM rate laws reveal, starvation and recovery rates are not independent parameters, and the biologically relevant portion of the phase space shown in Fig.~\ref{fig:fp} is constrained via covarying parameters.  
Given the parameters of terrestrial endotherms, we find that the starvation rate $\sigma$ and the recovery rate $\rho$ are constrained to lie within a small window of potential values (Fig.~\ref{fig:hopf}) for the known range of body sizes $M$. 
We thus find that the dynamics for all mammalian body sizes is confined to the steady-state regime of the NSM and that limit-cycle behavior is precluded.
Moreover, for larger $M$, the distance to the Hopf bifurcation increases, while uncertainty in allometric parameters (20\% variation around the mean; Fig.~\ref{fig:hopf}) results in little qualitative difference in the distance to the the Hopf bifurcation. 
These results suggest that small mammals are more prone to population oscillations---both stable limit cycles and transient cycles---than mammals with larger body size.  
Thus our NSM model predicts that population cycles should be less common for larger species and more common for smaller species, particularly in environments where resources are limiting.

\begin{figure}
\includegraphics[width=0.425\textwidth]{fig_DataHopf.pdf}
\caption{\small Transcritical (vertical lines) and Hopf bifurcations (curves) for
  allometrically determined starvation $\sigma$ and recovery $\rho$ rates as
  a function of different mammalian body sizes: $M=A\times10^1$g (blue) and
  $M=A\times10^6$g (green), where $A$ is a random uniform variable in $[1,9]$.
  Points denote realized values of $\sigma$ and $\rho$ given the drawn values for $M$.
  }
\label{fig:hopf}
\end{figure}

It should be noted that previous studies have used allometric constraints to
explain the periodicity of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$.  However this relation seems to hold only for some
species \cite{Hendriks:2012fc}, and potential drivers range from predator
and/or prey lifespans to competitive
dynamics~\cite{Kendall:1999iy,Hogstedt:2005cr}.  Statistically significant
support for the existence of population cycles among mammals is predominantly
based on time series for small mammals~\cite{Kendall:1998hl}, in agreement
with our predictions of more pronounced transient dynamics, given how close
these points are to the Hopf bifurcation.  On the other hand, the longer
gestational times and the increased difficulty in measurements, precludes
obtaining similar-quality data for larger organisms.

\section*{Extinction risk}
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  From the perspective of population
survival, it is the rate of starvation relative to the rate of recovery that
determines the long-term dynamics of the various species (Fig.~\ref{fig:fp}).
We therefore examine the competing effects of cyclic dynamics vs.\ changes in
steady state density on extinction risk as a function of $\sigma$ and $\rho$.
To this end, we computed the probability of extinction, where we define
extinction as a population trajectory falling below one tenth of the
allometrically constrained steady state at any time between $t=10^5$ and
$t \leq 10^8$.  This procedure is repeated for 1000 replicates of the
continuous-time system shown in Eq.~\ref{eq:system} for an organism of
$M=100$ grams.  In each replicate the initial densities are chosen to be
$A(F^*,H^*,R^*)$, with $A$ a random variable that is uniformly distributed in
$[0,2]$.  By allowing the rate of starvation to vary, we assessed extinction
risk across a range of values for $\sigma$ and $\rho$ between ca.\ $10^{-6}$
to $10^{-1}$.  As expected, higher rates of extinction correlate with both
high values of $\sigma$ if $\rho$ is small, and high values of $\rho$ if
$\sigma$ is small.  For low values of $\sigma$ and high values of $\rho$, the
increased extinction risk results from transient cycles with larger
amplitudes as the system nears the Hopf bifurcation (Fig.~\ref{fig:ext}).
For high values of $\sigma$ and low values of $\rho$, higher extinction risk
arises because of the decrease in the steady state consumer population
density (Figs.~\ref{fig:fp}\emph{B}, \ref{fig:ext}).  This interplay creates
an `extinction refuge', such that for a constrained range of $\sigma$ and
$\rho$, extinction probabilities are minimized.

\begin{figure}
\includegraphics[width=0.425\textwidth]{fig_ExtinctionAllometric.jpg}
\caption{\small Probability of extinction for a 100g consumer as a function of the
  starvation rate $\sigma$ and recovery rate $\rho$, where the initial
  density is given as $A(F^*,H^*,R^*)$, with $A$ being a random uniform
  variable in $[0,2]$.  Extinction is defined as the population trajectory
  falling below $0.1~\times$ the allometrically constrained steady state. The
  white point denotes the allometrically constrained starvation and recovery
  rate.}
\label{fig:ext}
\end{figure}

We find that the allometrically constrained values of $\sigma$ and $\rho$
fall squarely within the extinction refuge (Fig. \ref{fig:ext}, white point).
These values are close enough to the Hopf bifurcation to avoid low steady
state densities, and far enough away to avoid large-amplitude transient
cycles.  The fact that allometric values of $\sigma$ and $\rho$ fall within
this relatively small window supports the possibility that a selective
mechanism has constrained the physiological conditions that drive starvation
and recovery rates within populations.  Such a mechanism would select for
organism physiology that generates appropriate $\sigma$ and $\rho$ values
that serve to minimize extinction risk.  This selection could occur via the
tuning of body fat percentages, metabolic rates, and biomass maintenance
efficiencies.  To summarize, our finding that the allometrically-determined
parameters fall within this low extinction probability region suggests that
the NSM dynamics may both drive---and constrain---natural animal populations.

\section*{Dynamic and energetic barriers to body size}
Metabolite transport constraints are widely thought to place strict
boundaries on biological
scaling~\cite{Brown:1993p708,West:1997cg,Brown:2004wq} and thereby lead to
specific predictions on the minimum possible body size for
organisms~\cite{West:2002ud}.  Above this bound, a number of energetic and
evolutionary mechanisms have been explored to assess the costs and benefits
associated with larger body masses, particularly for mammals.  One important
such example is the \emph{fasting endurance hypothesis}, which contends that
larger body size, with consequent lower metabolic rates and increased ability
to maintain more endogenous energetic reserves, may buffer organisms against
environmental fluctuations in resource availability~\cite{Millar:1990p923}.
Over evolutionary time, terrestrial mammalian lineages show a significant
trend towards larger body size (known as Cope's
rule)~\cite{Alroy:1998p1594,Clauset:2009fh,Smith:2010p3442,Saarinen:2014br},
and it is thought that within-lineage drivers generate selection towards an
optimal upper bound of roughly $10^7$ grams~\cite{Alroy:1998p1594}, a value
that is likely limited by higher extinction risk for large taxa over longer
timescales~\cite{Clauset:2009fh}.  These trends are thought to be driven by a
combination of climate change and niche availability~\cite{Saarinen:2014br};
however the underpinning energetic costs and benefits of larger body sizes,
and how they influence dynamics over ecological timescales, have not been
explored.  We argue that the NSM provides a suitable framework to explore
these issues.

\begin{figure}
\includegraphics[width=0.4\textwidth]{fig_FPAllometric.pdf}
\caption{\small (\emph{A}) Consumer steady states $F^*$ (green) and $H^*$
  (orange) as a function of body mass.  (\emph{B}) Total energetic use
  $B_{\rm tot}$ of consumer populations at the steady state as a function of
  body mass.  (\emph{C}) Resource steady state $R^*$ as a function of
  consumer body mass.}
\label{fig:mass}
\end{figure}

@@

To examine whether the NSM could provide such a mechanism, we begin by noting
that a theoretical upper bound on mammalian body size is given by
$\epsilon_\sigma=0$, where mammals are entirely composed of metabolic
reserves, and this occurs at $M=8.3\times 10^8$, or $4.5\times$ the mass of a
blue whale.  Next we examine to what extent a more realistic upper bound to
body mass may serve as an evolutionary attractor, thus providing a suitable
within-lineage mechanism for Cope's rule.  We directly assess the
susceptibility of an otherwise homogeneous population to invasion by a
mutated subset of the population (denoted by ${}^\prime$) where individuals
have a modified proportion of body fat $M^\prime=M(1+\chi)$ where
$\chi \in [-1,1]$, thus altering the rates of starvation $\sigma(M^\prime)$,
recovery $\rho(M^\prime)$, and maintenance $\beta(M^\prime)$.  There is no
internal fixed point corresponding to a state where both original residents
and invaders coexist (except for the trivial state $\chi=0$).  To assess the
susceptibility to invasion as a function of the invader mass, we determine
which consumer has a higher steady-state density for a given value of $\chi$.
We find that for $1\leq M<8.43\times 10^6$g, having additional body fat
($\chi > 0$) results in a higher steady-state invader population density
($H^{\prime *}+F^{\prime *}>H^*+F^*$).  Thus the invader has an intrinsic
advantage over the resident population.  However, for $M>8.43\times 10^6$,
leaner individuals ($\chi < 0$) have advantageous steady state densities
(Fig.~\ref{fig:invasion}).
